---
title: "PFAS Analysis"
author: "Elinne Becket"
output: rmdformats::readthedown
---

## =========================================================
## PFAS sediment metagenome analysis
## Input:
##   1) PFAS_reads_per_kb_ref_spike_scaled_per_mg_matrix.tsv
##   2) MB_sediment_metadata.txt
## =========================================================

# Loading Required Libraries 
```{r, warning=FALSE, message=FALSE}

# Load the packages we need
library(tidyverse)
library(vegan)
library(pheatmap)
library(RColorBrewer)
library(broom)

```

## -------------------------
## File paths
## -------------------------
```{r}

pfas_file <- "PFAS_reads_per_kb_ref_spike_scaled_per_mg_matrix.tsv"
meta_file <- "MB_sediment_metadata.txt"

outdir <- "PFAS_R_analysis"
dir.create(outdir, showWarnings = FALSE)

```

## -------------------------
## Read PFAS matrix
## Rows = references
## Cols = samples
## -------------------------
```{r}

# Read PFAS abundance matrix (genes x samples)
pfas_mat_df <- read.delim(pfas_file, check.names = FALSE)

# Check that first column contains gene IDs
stopifnot(colnames(pfas_mat_df)[1] == "reference")

# Convert to matrix with genes as rows
pfas_mat <- pfas_mat_df |>
  column_to_rownames("reference") |>
  as.matrix()

# Transpose so rows = samples, columns = genes (needed for most analyses)
pfas_sample_gene <- t(pfas_mat)

# Ensure values are numeric
mode(pfas_sample_gene) <- "numeric"

# Store sample IDs
sample_ids <- rownames(pfas_sample_gene)

```


## -------------------------
## Read metadata + reshaping
## -------------------------

```{r}
# Read metadata and create sample IDs that match PFAS matrix
meta2 <- read.delim(meta_file, check.names = FALSE) |>
  mutate(
    Month_Code = sprintf("%02d", as.integer(Month_Code)),
    sample_id = paste(if_else(Month_Code == "01", "26", "25"), Month_Code, Site_Code, sep = "_"),
    Month = factor(Month, levels = c("October", "November", "December", "January"))
  ) |>
  filter(sample_id %in% sample_ids) |>
  slice(match(sample_ids, sample_id))

# Check sample order matches PFAS matrix
stopifnot(all(meta2$sample_id == rownames(pfas_sample_gene)))

# Convert PFAS matrix to long format for plotting
long_df <- as.data.frame(pfas_sample_gene) |>
  tibble::rownames_to_column("sample_id") |>
  tidyr::pivot_longer(-sample_id, names_to = "reference", values_to = "abundance") |>
  left_join(meta2, by = "sample_id")

```

## -------------------------
## Site ordering
## -------------------------
```{r}
# Set consistent site order for all plots
site_order <- c(
  "Mission Point", "Fanuel St Park", "Paradise Point", "Fiesta Sunset Beach",
  "De Anza Cove", "South Shores", "Leisure Lagoon", "Tecolote Creek"
)

long_df$Site <- factor(long_df$Site, levels = site_order)
meta2$Site <- factor(meta2$Site, levels = site_order)

```


## -------------------------
## Add summary columns and transform data for plotting
## -------------------------

```{r}
# Calculate total PFAS signal per sample and per gene
sample_sums <- rowSums(pfas_sample_gene, na.rm = TRUE)
gene_sums   <- colSums(pfas_sample_gene, na.rm = TRUE)

# Add summary metrics to metadata
alpha_df <- meta2 |>
  mutate(
    total_PFAS_abundance = sample_sums,
    richness_detected = rowSums(pfas_sample_gene > 0, na.rm = TRUE),
    shannon = vegan::diversity(pfas_sample_gene, index = "shannon")
  )

# Log-transform data to reduce influence of large values
pseudo <- min(pfas_sample_gene[pfas_sample_gene > 0], na.rm = TRUE) / 2
pfas_log <- log10(pfas_sample_gene + pseudo)
```


### Analyses


## -------------------------
## 1) Boxplots
## -------------------------

```{r,warning = FALSE,message=FALSE,fig.width= 15, fig.height=8}

# Sum all PFAS genes within each sample
sample_totals <- long_df |>
  group_by(sample_id, Site) |>
  summarise(total_abundance = sum(abundance, na.rm = TRUE), .groups = "drop")

# Plot total PFAS abundance across sites
p_box_total <- ggplot(sample_totals, aes(x = Site, y = total_abundance, fill = Site)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(width = 0.15, size = 2, alpha = 0.8) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(
    title = "Total PFAS Gene Abundance by Site",
    x = "Site",
    y = "PFAS abundance (spike-scaled per mg sediment)"
  )

# Save and show the plot
ggsave("plots/Total PFAS Gene Abundance by Site.pdf", p_box_total, width = 8, height = 8, limitsize = FALSE)
p_box_total

# Plot abundance of each PFAS gene across sites
p_gene_box <- ggplot(long_df, aes(x = Site, y = abundance)) +
  geom_boxplot(outlier.shape = NA, fill = "grey80") +
  geom_jitter(width = 0.15, size = 1, alpha = 0.6) +
  facet_wrap(~ reference, scales = "free_y") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(
    title = "PFAS Gene Abundance by Site (All Genes)",
    x = "Site",
    y = "PFAS abundance (spike-scaled per mg sediment)"
  )

# Save and show the faceted gene plot
ggsave("plots/PFAS Gene Abundance by Site.pdf", p_gene_box, width = 18, height = 12, limitsize = FALSE)
p_gene_box


```

## -------------------------
## 2) Top genes heatmap
## -------------------------

```{r, fig.width= 10, fig.height=10}

# Select the top 20 most abundant PFAS genes
top_n <- 20
top_genes <- names(sort(gene_sums, decreasing = TRUE))[1:min(top_n, ncol(pfas_log))]

# Add sample metadata for the heatmap columns
ann_col <- meta2 |>
  select(sample_id, Month, Site) |>
  column_to_rownames("sample_id")

# Subset to top genes and transpose so genes are rows
heat_mat <- t(pfas_log[, top_genes, drop = FALSE])

# Plot heatmap of relative abundance patterns
hm <- pheatmap(
  heat_mat,
  annotation_col = ann_col,
  scale = "row",
  clustering_distance_cols = "euclidean",
  clustering_method = "complete",
  fontsize_row = 8,
  fontsize_col = 9,
  border_color = NA,
  color = colorRampPalette(c("#2c7bb6", "white", "#d7191c"))(100),
  breaks = seq(-3, 3, length.out = 101)
)

# Save the heatmap
ggsave("plots/PFAS Gene Heatmap by Site and Month.pdf", hm, width = 10, height = 12, limitsize = FALSE)


```


## -------------------------
## 6) Correlation of total PFAS abundance with chemistry
## -------------------------
```{r}

# Environmental variables to test against total PFAS abundance
corr_vars <- c("Density", "pH", "sH2O", "Nperc", "Cperc", "C:N_ratio", "TC", "TN", "SOM",
               "NH4_mgkg", "NO3_mgkg", "DIN_mgkg", "PO4_mgkg")

# Run Spearman correlations for each variable
corr_results <- map_dfr(corr_vars, function(v) {
  x <- alpha_df[[v]]
  y <- alpha_df$total_PFAS_abundance
  ct <- suppressWarnings(cor.test(x, y, method = "spearman"))
  tibble(
    variable = v,
    rho = unname(ct$estimate),
    p_value = ct$p.value
  )
}) |>
  mutate(p_adj = p.adjust(p_value, method = "BH")) |>
  arrange(p_adj, p_value)

# Format results for plotting
corr_plot_df <- corr_results |>
  mutate(
    variable = factor(variable, levels = rev(variable)),
    sig = if_else(p_adj < 0.05, "FDR < 0.05", "ns"),
    direction = if_else(rho > 0, "Positive", "Negative")
  )

# Plot correlation strength for each variable
p_corr <- ggplot(corr_plot_df, aes(x = rho, y = variable, color = sig, shape = direction)) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey60") +
  geom_point(size = 3) +
  theme_bw() +
  labs(
    title = "Correlation of Total PFAS Abundance with Sediment Chemistry",
    x = "Spearman rho",
    y = "Environmental variable",
    color = NULL
  )

# Save and show the plot
ggsave("plots/Correlation of Total PFAS Abundance with Sediment Chemistry.pdf", p_corr, width = 8, height = 4, limitsize = FALSE)
p_corr

```


## -------------------------
## 7) Gene-wise associations with metadata
## Spearman against selected variables
## -------------------------
```{r}

# Environmental variables to test against each PFAS gene
selected_vars <- c("Density", "pH", "sH2O", "Nperc", "Cperc", "C:N_ratio", "TC", "TN", "SOM",
               "NH4_mgkg", "NO3_mgkg", "DIN_mgkg", "PO4_mgkg")

# Run Spearman correlations for each gene-variable pair
gene_cor_results <- map_dfr(colnames(pfas_sample_gene), function(g) {
  y <- pfas_sample_gene[, g]

  map_dfr(selected_vars, function(v) {
    x <- meta2[[v]]
    ct <- suppressWarnings(cor.test(x, y, method = "spearman"))
    tibble(
      reference = g,
      variable = v,
      rho = unname(ct$estimate),
      p_value = ct$p.value
    )
  })
}) |>
  mutate(p_adj = p.adjust(p_value, method = "BH")) |>
  arrange(p_adj, p_value)

# Convert correlation results into a matrix for plotting
gene_cor_mat <- gene_cor_results |>
  select(reference, variable, rho) |>
  pivot_wider(names_from = variable, values_from = rho) |>
  column_to_rownames("reference") |>
  as.matrix()

# Mark significant correlations with an asterisk
sig_mat <- gene_cor_results |>
  mutate(sig = if_else(p_adj < 0.05, "*", "")) |>
  select(reference, variable, sig) |>
  pivot_wider(names_from = variable, values_from = sig) |>
  column_to_rownames("reference") |>
  as.matrix()

# Plot heatmap of gene-environment correlations
hm_cor <- pheatmap(
  gene_cor_mat,
  color = colorRampPalette(c("#2c7bb6", "white", "#d7191c"))(100),
  breaks = seq(-1, 1, length.out = 101),
  border_color = NA,
  fontsize_row = 7,
  fontsize_col = 10,
  display_numbers = sig_mat,
  number_color = "black",
  main = "Spearman Correlation of PFAS Genes with Environmental Variables"
)

# Save the heatmap
ggsave("plots/Spearman Correlation of PFAS Genes with Environmental Variables.pdf", hm_cor, width = 8, height = 6, limitsize = FALSE)


```




